---
title: "random_forest"
author: "Leonardo Feitosa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
library(tidyverse)
library(tidymodels)
library(here)
library(randomForest)
library(ranger)
library(googlesheets4)
library(naniar)
library(vip)

set.seed(42)
```

```{r}
prm_elasmo <- read_csv(here("data", "prm_hypoxia.csv")) %>%
  filter(gear_class == "longline") %>%
  filter(!grepl("Rajidae", family)) %>%
  filter(family != "Dasyatidae") %>%
  filter(family != "Pristidae") %>%
  filter(family != "Rhinidae") %>%
  filter(family != "Rhinobatidae") %>%
  mutate(habitat = case_when(
    habitat_associated == "pelagic" ~ "pelagic",
    habitat_associated == "reef-associated" ~ "pelagic",
    habitat_associated == "bathypelagic" ~ "pelagic",
    habitat_associated == "benthopelagic" ~ "pelagic",
    habitat_associated == "demersal" ~ "demersal",
    habitat_associated == "bathydemersal" ~ "demersal",
  )) %>%
  mutate(ventilation_method = case_when(
    scientific_name == "Ginglymostoma cirratum" ~ "stationary",
    TRUE ~ ventilation_method
  )) %>%
  mutate(reproductive_mode = case_when(
    scientific_name == "Hexanchus nakamurai" ~ "yolk-sac viviparity",
    TRUE ~ reproductive_mode
  )) %>%
  group_by(scientific_name) %>%
  mutate(median_depth = mean(median_depth)) %>%
  distinct()
```

```{r}
# create data subset with variables that will be used in the random forest model
prm_elasmo_subset <- prm_elasmo %>%
  select(
    scientific_name, gear_class, habitat, estimate_type, estimate, sample_size,
    method, ventilation_method, max_size_cm, ac, measure, median_depth, evol_dist, reproductive_mode, family
  ) %>%
  mutate(measure = case_when(
    scientific_name == "sharks" ~ "total_length",
    scientific_name %in% c("rays", "Dasyatis sp") ~ "disk_width",
    TRUE ~ measure
  )) %>%
  mutate(reproductive_mode = case_when(
    scientific_name == "sharks" ~ "yolk-sac viviparity",
    scientific_name == "rays" ~ "yolk-sac viviparity",
    TRUE ~ reproductive_mode
  )) %>%
  mutate_if(is.character, as.factor) %>%
  filter(!grepl("sp", scientific_name))

# Create subset of the whole dataset with at-vessel mortality and calculate the weighted averages per species and gear
elasmo_avm <- prm_elasmo_subset %>%
  filter(estimate_type == "at-vessel mortality") %>%
  filter(sample_size > 5) %>%
  group_by(scientific_name, gear_class) %>%
  mutate(mortality_prop = weighted.mean(estimate, sample_size)) %>%
  select(ventilation_method, median_depth, max_size_cm, habitat, reproductive_mode, evol_dist, ac, mortality_prop, family) %>%
  distinct() %>%
  drop_na()

# Create subset of the whole dataset with post-release mortality and calculate the weighted averages per species and gear
elasmo_prm <- prm_elasmo_subset %>%
  filter(estimate_type == "post-release mortality") %>%
  filter(sample_size > 5) %>%
  group_by(scientific_name, gear_class) %>%
  mutate(mortality_prop = weighted.mean(estimate, sample_size)) %>%
  select(ventilation_method, median_depth, max_size_cm, habitat, reproductive_mode, evol_dist, mortality_prop, family) %>%
  distinct() %>%
  drop_na()
```

```{r}
# split the data into training and test set
## At-vessel mortality
avm_split <- initial_split(elasmo_avm, prop = .7, strata = family)

avm_train <- training(avm_split)
avm_test <- testing(avm_split)

avm_folds <- vfold_cv(avm_train, repeats = 5, v = 10, strata = family)

## Post-release mortality
# prm_split <- initial_split(elasmo_prm, prop = .7)
#
# prm_train <- training(prm_split)
# prm_test <- testing(prm_split)

prm_folds <- vfold_cv(elasmo_prm, repeats = 5, v = 10, strata = family)
```


## Overview

This model aims to predict the likelihood of mortality per taxonomic family between sharks and rays using several species-specific bioecological features such as ventilation mode, median temperature, reproductive mode, median depth, as well as technological features such as the fishing gear each species was captured with.

```{r}
# Create the recipes for each estimate
avm_recipe <- recipe(mortality_prop ~ median_depth + max_size_cm +  reproductive_mode + evol_dist + ac, data = avm_train) %>%
  step_normalize(all_numeric_predictors())

prm_recipe <- recipe(mortality_prop ~ median_depth + max_size_cm +  reproductive_mode + evol_dist, data = elasmo_prm) %>%
  step_normalize(all_numeric_predictors())
```

```{r}
# create models
avm_rf <- rand_forest(trees = tune(), mtry = tune(), min_n = tune(), mode = "regression") %>%
  set_engine("ranger", verbose = TRUE, oob.error = TRUE, quantreg = TRUE, splitrule = "beta", importance = "impurity")

prm_rf <- rand_forest(trees = tune(), mtry = tune(), min_n = tune(), mode = "regression") %>%
  set_engine("ranger", verbose = TRUE, oob.error = TRUE, quantreg = TRUE, splitrule = "beta", importance = "impurity")
```

```{r}
# create workflows
avm_wflw <- workflow() %>%
  add_model(avm_rf) %>%
  add_recipe(avm_recipe)

prm_wflw <- workflow() %>%
  add_model(prm_rf) %>%
  add_recipe(prm_recipe)
```

### Cross Validation

Cross validation will be used to choose the best hyperparameters for use in the model: 

* Number of trees
* Number of predictors in each sample
* Minimum data points in each node required for node split

```{r,include=FALSE}
avm_tune <- avm_wflw %>% # use cross validation to tune parameters
  tune_grid(
    resamples = avm_folds, # add cv folds
    grid = 10 # grid_regular(trees(), finalize(mtry(), avm_train), min_n(), levels = 10) %>% filter(mtry < 7)
  ) # use grid of parameters to test
```

```{r,include=FALSE}
prm_tune <- prm_wflw %>% # use cross validation to tune parameters
  tune_grid(
    resamples = prm_folds, # add cv folds
    grid = 10# grid_regular(trees(), finalize(mtry(), elasmo_prm), min_n(), levels = 10) %>% filter(mtry < 7)
  ) # use grid of parameters to test
```

Below displays the results of cross validation tuning

```{r}
# autoplot(avm_tune) + # plot results of tuning
#   theme_bw() # change theme
```

```{r}
select_best(avm_tune, metric = "rsq", n = 1) # get the best model
```

```{r, include=FALSE}
avm_final <- finalize_workflow(avm_wflw, select_best(avm_tune, metric = "rsq")) # finalize the workflow with the best model
#
# saveRDS(avm_final, here::here("data", "avm_final.rds"))

# avm_final <- readRDS(here::here("data", "avm_final.rds"))
```

```{r, include=FALSE}
avm_fit <- fit(avm_final, avm_train) # fit the data to the training data
```

```{r, include=FALSE}
avm_train_predict <- predict(object = avm_fit, new_data = avm_train) %>% # predict the training set
  bind_cols(avm_train) # bind training set column to prediction

avm_test_predict <- predict(object = avm_fit, new_data = avm_test) %>% # predict the training set
  bind_cols(avm_test) # bind prediction to testing data column
```

```{r}
avm_train_metrics <- avm_train_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
avm_train_metrics

avm_test_metrics <- avm_test_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
avm_test_metrics
```

Below displays the results of cross validation tuning

```{r}
# autoplot(prm_tune) + # plot results of tuning
#   theme_bw() # change theme
```

```{r}
# select_best(prm_tune, metric = "rsq", n = 1) # get the best model
```

```{r, include=FALSE}
prm_final <- finalize_workflow(prm_wflw, select_best(prm_tune, metric = "rsq")) # finalize the workflow with the best model
#
# saveRDS(prm_final, here::here("data", "prm_final.rds"))

# prm_final <- readRDS(here::here("data", "prm_final.rds"))
```

```{r, include=FALSE}
prm_fit <- fit(prm_final, elasmo_prm) # fit the data to the training data
```

```{r, include=FALSE}
prm_train_predict <- predict(object = prm_fit, new_data = elasmo_prm) %>% # predict the training set
  bind_cols(elasmo_prm) # bind training set column to prediction

# prm_test_predict <- predict(object = prm_fit, new_data = prm_test) %>% # predict the training set
#   bind_cols(prm_test)# bind prediction to testing data column
```

```{r}
prm_train_metrics <- prm_train_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
prm_train_metrics

# prm_test_metrics <- prm_test_predict %>%
#   metrics(mortality_prop, .pred) # get testing data metrics
# prm_test_metrics
```

```{r}
my_controls <- control_resamples(
  verbose = TRUE,
  save_pred = TRUE,
  extract = function(x) {
    model <- extract_fit_engine(x)
    model$prediction.error
  }
) # set to get prediction and out of bag errors

avm_cv2 <- avm_final %>%
  fit_resamples(avm_folds, control = my_controls) # fit the folds to the final model

prm_cv2 <- prm_final %>%
  fit_resamples(prm_folds, control = my_controls) # fit the folds to the final model
```

```{r}
collect_metrics(avm_cv2)

collect_metrics(prm_cv2)
```

### Variable importance for the at-vessel mortality and post-release mortality RF models

```{r}
avm_fit %>%
  extract_fit_parsnip() %>%
  vip()

prm_fit %>%
  extract_fit_parsnip() %>%
  vip()
```
