---
title: "random_forest"
author: "Leonardo Feitosa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
library(tidyverse)
library(tidymodels)
library(here)
library(randomForest)
library(ranger)
library(googlesheets4)
library(naniar)
library(vip)

# Set data paths
basedir <- "G:/Meu Drive/PRM review/"
datadir <- file.path(basedir, "data/fish_base_data")
outdir <- file.path(basedir, "data/outputs") # calls present data from raster files

set.seed(42)
```

```{r}
prm_elasmo <- read_csv(here("data", "prm_dataset.csv")) 

prm_elasmo <- prm_elasmo %>%
  rename(scientific_name = species)

# sp counts per estimate type
avm_sp_count <- prm_elasmo %>%
  filter(estimate_type == "at-vessel mortality") %>%
  distinct(scientific_name) %>%
  count()

prm_sp_count <- prm_elasmo %>%
  filter(estimate_type == "post-release mortality") %>%
  distinct(scientific_name) %>%
  count()
```

```{r}
# create data subset with variables that will be used in the random forest model
prm_elasmo_subset <- prm_elasmo %>%
  select(
    scientific_name, gear_class, habitat_associated, estimate_type, estimate, sample_size,
    method, ventilation_method, max_size_cm, measure, median_depth, reproductive_mode, family
  ) %>%
   mutate(max_size_cm = case_when( # filling missing information based on the average per group
    scientific_name == "Dasyatis sp" ~ 146.22,
    scientific_name == "Squalus sp" ~ 141.12,
    scientific_name == "Mustelus sp" ~ 171.9,
    scientific_name == "Centrophorus sp" ~ 165.75,  
    scientific_name == "sharks" ~ 347.2639,
    scientific_name == "rays" ~ 170.2222, # only calculated for the batoid species with disk width measures
    TRUE ~ max_size_cm
  )) %>% 
  mutate(median_depth = case_when( # filling missing information based on the average per group
    scientific_name == "Dasyatis sp" ~ 139.33,
    scientific_name == "Squalus sp" ~ 601.875,
    scientific_name == "Mustelus sp" ~ 254.3929,
    scientific_name == "Centrophorus sp" ~ 1136.25,
    scientific_name == "sharks" ~ 313.2104,
    scientific_name == "rays" ~ 120.5278,
    TRUE ~ median_depth
  )) %>%
  mutate(measure = case_when(
    scientific_name == "sharks" ~ "total_length",
    scientific_name %in% c("rays", "Dasyatis sp") ~ "disk_width",
    TRUE ~ measure
  )) %>%
  mutate(reproductive_mode = case_when(
    scientific_name == "sharks" ~ "yolk-sac viviparity",
    scientific_name == "rays" ~ "yolk-sac viviparity",
    TRUE ~ reproductive_mode
  )) %>%
  mutate_if(is.character, as.factor)

# Create subset of the whole dataset with at-vessel mortality and calculate the weighted averages per species and gear
elasmo_trawl <- prm_elasmo_subset %>%
  filter(estimate_type == "at-vessel mortality" & estimate > 0) %>%
  filter(gear_class == "trawl") %>% 
  mutate(est_count = sample_size * estimate) %>%
  group_by(scientific_name, gear_class) %>%
  mutate(mortality_prop = sum(est_count) / sum(sample_size)) %>%
  drop_na()

# Create subset of the whole dataset with post-release mortality and calculate the weighted averages per species and gear
elasmo_gillnet <- prm_elasmo_subset %>%
  filter(estimate_type == "at-vessel mortality" & estimate > 0) %>%
  filter(gear_class == "gillnet") %>% 
  mutate(est_count = sample_size * estimate) %>%
  group_by(scientific_name, gear_class) %>%
  mutate(mortality_prop = sum(est_count) / sum(sample_size)) %>%
  drop_na()
```


```{r}
# split the data into training and test set

trawl_folds <- vfold_cv(elasmo_trawl, repeats = 5, v = 10)

gillnet_folds <- vfold_cv(elasmo_gillnet, repeats = 5, v = 10)
```


## Overview

This model aims to predict the likelihood of mortality per taxonomic family between sharks and rays using several species-specific bioecological features such as ventilation mode, median temperature, reproductive mode, median depth, as well as technological features such as the fishing gear each species was captured with.

```{r}
# Create the recipes for each estimate
trawl_recipe <- recipe(mortality_prop ~ median_depth + max_size_cm + habitat_associated + reproductive_mode, data = elasmo_trawl) %>%
  step_normalize(all_numeric_predictors())

gillnet_recipe <- recipe(mortality_prop ~ median_depth + max_size_cm + habitat_associated + reproductive_mode, data = elasmo_gillnet) %>%
  step_normalize(all_numeric_predictors())
```

```{r}
# create models
trawl_rf <- rand_forest(trees = tune(), mtry = tune(), min_n = tune(), mode = "regression") %>%
  set_engine("ranger", verbose = TRUE, oob.error = TRUE, quantreg = TRUE, splitrule = "beta", importance = "impurity")

gillnet_rf <- rand_forest(trees = tune(), mtry = tune(), min_n = tune(), mode = "regression") %>%
  set_engine("ranger", verbose = TRUE, oob.error = TRUE, quantreg = TRUE, splitrule = "beta", importance = "impurity")
```

```{r}
# create workflows
trawl_wflw <- workflow() %>%
  add_model(trawl_rf) %>%
  add_recipe(trawl_recipe)

gillnet_wflw <- workflow() %>%
  add_model(gillnet_rf) %>%
  add_recipe(gillnet_recipe)
```

### Cross Validation

Cross validation will be used to choose the best hyperparameters for use in the model: 

* Number of trees
* Number of predictors in each sample
* Minimum data points in each node required for node split

```{r,include=FALSE}
trawl_tune <- trawl_wflw %>% # use cross validation to tune parameters
  tune_grid(
    resamples = trawl_folds, # add cv folds
    grid = 10
  ) # use grid of parameters to test
```

```{r,include=FALSE}
gillnet_tune <- gillnet_wflw %>% # use cross validation to tune parameters
  tune_grid(
    resamples = gillnet_folds, # add cv folds
    grid = 10
  ) # use grid of parameters to test
```

Below displays the results of cross validation tuning

```{r}
autoplot(trawl_tune) + # plot results of tuning
  theme_bw() # change theme
```

```{r}
select_best(trawl_tune, metric = "rsq", n = 1) # get the best model
```

```{r, include=FALSE}
trawl_final <- finalize_workflow(trawl_wflw, select_best(trawl_tune, metric = "rsq")) # finalize the workflow with the best model
```

```{r, include=FALSE}
trawl_fit <- fit(trawl_final, elasmo_trawl) # fit the data to the training data
```

```{r, include=FALSE}
trawl_train_predict <- predict(object = trawl_fit, new_data = elasmo_trawl) %>% # predict the training set
  bind_cols(elasmo_trawl) # bind training set column to prediction
```

```{r}
trawl_train_metrics <- trawl_train_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
trawl_train_metrics
```

Below displays the results of cross validation tuning

```{r}
autoplot(gillnet_tune) + # plot results of tuning
  theme_bw() # change theme
```

```{r}
select_best(gillnet_tune, metric = "rsq", n = 1) # get the best model
```

```{r, include=FALSE}
gillnet_final <- finalize_workflow(gillnet_wflw, select_best(gillnet_tune, metric = "rsq")) # finalize the workflow with the best model
```

```{r, include=FALSE}
gillnet_fit <- fit(gillnet_final, elasmo_gillnet) # fit the data to the training data
```

```{r, include=FALSE}
gillnet_train_predict <- predict(object = gillnet_fit, new_data = elasmo_gillnet) %>% # predict the training set
  bind_cols(elasmo_gillnet) # bind training set column to prediction
```

```{r}
gill_train_metrics <- gillnet_train_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
gill_train_metrics
```

```{r}
my_controls <- control_resamples(
  verbose = TRUE,
  save_pred = TRUE,
  extract = function(x) {
    model <- extract_fit_engine(x)
    model$prediction.error
  }
) # set to get prediction and out of bag errors

trawl_cv2 <- trawl_final %>%
  fit_resamples(trawl_folds, control = my_controls) # fit the folds to the final model

gillnet_cv2 <- gillnet_final %>%
  fit_resamples(gillnet_folds, control = my_controls) # fit the folds to the final model
```

```{r}
collect_metrics(trawl_cv2)

collect_metrics(gillnet_cv2)
```

