---
title: "random_forest"
author: "Leonardo Feitosa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
library(tidyverse)
library(tidymodels)
library(here)
library(ranger)
library(naniar)
library(vip)
library(patchwork)
library(ggpmisc)

set.seed(42)
```

```{r}
# read in the data
prm_elasmo <- read_csv(here("data", "prm_elasmo_subset.csv")) %>%
  filter(gear_class == "longline") %>%
  filter(!grepl("Rajidae", family)) %>%
  filter(family != "Dasyatidae") %>%
  filter(family != "Pristidae") %>%
  filter(family != "Rhinidae") %>%
  filter(family != "Rhinobatidae")
```

```{r}
# get only PRM data
prm <- prm_elasmo %>%
  filter(estimate_type == "post-release mortality") %>%
  rename(prm = estimate) %>%
  select(scientific_name, prm)

# get only AVM data
avm <- prm_elasmo %>%
  filter(estimate_type == "at-vessel mortality") %>%
  rename(avm = estimate) %>%
  select(scientific_name, avm)
```

```{r}
# correlation between avm and prm
cor_check <- full_join(avm, prm) %>%
  filter(!is.na(avm) & !is.na(prm))

cor(cor_check$prm, cor_check$avm)
```

```{r}
# create data subset with variables that will be used in the random forest model
prm_elasmo_subset <- prm_elasmo %>%
  select(
    scientific_name, gear_class, habitat_associated, estimate_type, estimate, sample_size,
    method, ventilation_method, max_size_cm, measure, median_depth, reproductive_mode, family
  ) %>%
  mutate(measure = case_when(
    scientific_name == "sharks" ~ "total_length",
    scientific_name %in% c("rays", "Dasyatis sp") ~ "disk_width",
    TRUE ~ measure
  )) %>%
  mutate(reproductive_mode = case_when(
    scientific_name == "sharks" ~ "yolk-sac viviparity",
    scientific_name == "rays" ~ "yolk-sac viviparity",
    TRUE ~ reproductive_mode
  )) %>%
  mutate_if(is.character, as.factor) %>%
  filter(!grepl("sp", scientific_name))

# Create subset of the whole dataset with at-vessel mortality and calculate the weighted averages per species and gear
elasmo_avm <- prm_elasmo_subset %>%
  filter(estimate_type == "at-vessel mortality" & estimate > 0) %>%
  mutate(est_count = sample_size * estimate) %>%
  group_by(scientific_name, gear_class) %>%
  mutate(mortality_prop = sum(est_count) / sum(sample_size)) %>%
  select(-sample_size, -estimate, -est_count) %>%
  distinct() %>%
  drop_na()

# Create subset of the whole dataset with post-release mortality and calculate the weighted averages per species and gear
elasmo_prm <- prm_elasmo_subset %>%
  filter(estimate_type == "post-release mortality" & estimate > 0) %>%
  mutate(est_count = sample_size * estimate) %>%
  group_by(scientific_name, gear_class) %>%
  mutate(mortality_prop = sum(est_count) / sum(sample_size)) %>%
  select(-sample_size, -estimate, -est_count) %>%
  distinct() %>%
  drop_na()
```

```{r}
# split the data into training and test set for At-vessel mortality
avm_split <- initial_split(elasmo_avm, prop = .8)

avm_train <- training(avm_split)
avm_test <- testing(avm_split)

# cross validation sets
avm_folds <- vfold_cv(avm_train, repeats = 5, v = 10)

prm_folds <- vfold_cv(elasmo_prm, repeats = 5, v = 10)
```

## Overview

This model aims to predict the likelihood of mortality per taxonomic family between sharks and rays using several species-specific bioecological features such as ventilation mode, median temperature, reproductive mode, median depth, as well as technological features such as the fishing gear each species was captured with.

```{r}
# Create the recipes for each estimate
avm_recipe <- recipe(mortality_prop ~ reproductive_mode + median_depth + max_size_cm + habitat_associated + ventilation_method, data = avm_train) %>%
  step_normalize(all_numeric_predictors())

prm_recipe <- recipe(mortality_prop ~ reproductive_mode + max_size_cm + habitat_associated + ventilation_method, data = elasmo_prm) %>%
  step_normalize(all_numeric_predictors())
```

```{r}
# create models
avm_rf <- rand_forest(trees = tune(), mtry = tune(), min_n = tune(), mode = "regression") %>%
  set_engine("ranger", verbose = TRUE, oob.error = TRUE, quantreg = TRUE, splitrule = "beta", importance = "impurity")

prm_rf <- rand_forest(trees = tune(), mtry = tune(), min_n = tune(), mode = "regression") %>%
  set_engine("ranger", verbose = TRUE, oob.error = TRUE, quantreg = TRUE, splitrule = "beta", importance = "impurity")
```

```{r}
# create workflows
avm_wflw <- workflow() %>%
  add_model(avm_rf) %>%
  add_recipe(avm_recipe)

prm_wflw <- workflow() %>%
  add_model(prm_rf) %>%
  add_recipe(prm_recipe)
```

### Cross Validation

Cross validation will be used to choose the best hyperparameters for use in the model: 

* Number of trees
* Number of predictors in each sample
* Minimum data points in each node required for node split

*tuned data provided

```{r,include=FALSE}
# avm_tune <- avm_wflw %>% # use cross validation to tune parameters
#   tune_grid(
#     resamples = avm_folds, # add cv folds
#     grid = 10
#   ) # use grid of parameters to test
```

```{r,include=FALSE}
# prm_tune <- prm_wflw %>% # use cross validation to tune parameters
#   tune_grid(
#     resamples = prm_folds, # add cv folds
#     grid = 10
#   ) # use grid of parameters to test
```

Below displays the results of cross validation tuning for AVM

```{r}
# autoplot(avm_tune) + # plot results of tuning
#   theme_bw() # change theme
```

```{r}
# select_best(avm_tune, metric = "rsq", n = 1) # get the best model
```

```{r, include=FALSE}
# avm_final <- finalize_workflow(avm_wflw, select_best(avm_tune, metric = "rsq")) # finalize the workflow with the best model
#
# saveRDS(avm_final, here::here("data", "avm_final.rds"))

avm_final <- readRDS(here::here("data", "avm_final.rds"))
```

```{r, include=FALSE}
avm_fit <- fit(avm_final, avm_train) # fit the data to the training data
```

```{r, include=FALSE}
avm_train_predict <- predict(object = avm_fit, new_data = avm_train) %>% # predict the training set
  bind_cols(avm_train) # bind training set column to prediction

avm_test_predict <- predict(object = avm_fit, new_data = avm_test) %>% # predict the training set
  bind_cols(avm_test) # bind prediction to testing data column
```

```{r}
avm_train_metrics <- avm_train_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
avm_train_metrics

avm_test_metrics <- avm_test_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
avm_test_metrics
```

Below displays the results of cross validation tuning for PRM

```{r}
# autoplot(prm_tune) + # plot results of tuning
#   theme_bw() # change theme
```

```{r}
# select_best(prm_tune, metric = "rsq", n = 1) # get the best model
```

```{r, include=FALSE}
# prm_final <- finalize_workflow(prm_wflw, select_best(prm_tune, metric = "rsq")) # finalize the workflow with the best model
#
# saveRDS(prm_final, here::here("data", "prm_final.rds"))

prm_final <- readRDS(here::here("data", "prm_final.rds"))
```

```{r, include=FALSE}
prm_fit <- fit(prm_final, elasmo_prm) # fit the data to the training data
```

```{r, include=FALSE}
prm_train_predict <- predict(object = prm_fit, new_data = elasmo_prm) %>% # predict the training set
  bind_cols(elasmo_prm) # bind training set column to prediction
```

```{r}
prm_train_metrics <- prm_train_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
prm_train_metrics
```

```{r}
my_controls <- control_resamples(
  verbose = TRUE,
  save_pred = TRUE,
  extract = function(x) {
    model <- extract_fit_engine(x)
    model$prediction.error
  }
) # set to get prediction and out of bag errors

avm_cv2 <- avm_final %>%
  fit_resamples(avm_folds, control = my_controls) # fit the folds to the final model

prm_cv2 <- prm_final %>%
  fit_resamples(prm_folds, control = my_controls) # fit the folds to the final model
```

```{r}
collect_metrics(avm_cv2)

collect_metrics(prm_cv2)
```

## Quantile regression

For AVM

```{r}
preds_bind <- function(data_fit, rf_fit) {
  predict(
    rf_fit$fit$fit$fit,
    workflows::extract_recipe(rf_fit) %>% bake(data_fit),
    type = "quantiles",
    quantiles = c(.05, .25, .5, .75, .95)
  ) %>%
    with(predictions) %>%
    as_tibble() %>%
    set_names(paste0(".pred", c("_05", "_25", "_50", "_75", "_95"))) %>%
    bind_cols(data_fit)
} # function to preform and extract quantile estimates from random forest

avm_data <- full_join(avm_test_predict, avm_train_predict)

# predict quantiles
avm_quant_test <- preds_bind(avm_data, avm_fit) %>%
  mutate(
    range_50 = .pred_75 - .pred_25,
    range_95 = .pred_95 - .pred_05
  )

q1 <- ggplot(avm_quant_test) +
  geom_boxplot(aes(range_50)) +
  theme_bw(base_size = 14) +
  labs(x = "Distribution of IQR for AVM") +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
q1

write_csv(avm_quant_test, here::here("data", "avm_model_predictions.csv"))
```

For PRM

```{r}
# predict quantiles
prm_quant_test <- preds_bind(prm_train_predict, prm_fit) %>%
  mutate(
    range_50 = .pred_75 - .pred_25,
    range_95 = .pred_95 - .pred_05
  )

q2 <- ggplot(prm_quant_test) +
  geom_boxplot(aes(range_50)) +
  theme_bw(base_size = 14) +
  labs(x = "Distribution of IQR for PRM") +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
q2

write_csv(prm_quant_test, here::here("data", "prm_model_predictions.csv"))
```

## AVM model - real values vs predicted values 

```{r}
p1 <- ggplot(avm_quant_test, aes(x = mortality_prop, y = .pred_50)) + # plot ln of real versus ln of predicted
  geom_point() +
  stat_poly_line() +
  stat_poly_eq(use_label("eq")) +
  stat_poly_eq(label.y = 0.9) +
  theme_bw(base_size = 14) +
  labs(
    x = "Observed AVM Rate",
    y = "Predicted AVM Rate"
  )
p1

summary(lm(.pred_50 ~ mortality_prop, data = avm_quant_test))
```

## PRM model - real values vs predicted values

```{r}
p2 <- ggplot(prm_train_predict, aes(x = mortality_prop, y = .pred)) + # plot ln of real versus ln of predicted
  geom_point() +
  stat_poly_line() +
  stat_poly_eq(use_label("eq")) +
  stat_poly_eq(label.y = 0.9) +
  theme_bw(base_size = 14) +
  labs(
    x = "Observed PRM Rate",
    y = "Predicted PRM Rate"
  )
p2

summary(lm(.pred ~ mortality_prop, data = prm_train_predict))
```

### Variable importance for the at-vessel mortality and post-release mortality RF models

```{r}
v1 <- avm_fit %>%
  extract_fit_parsnip() %>%
  vip() +
  theme_bw(base_size = 14) +
  labs(y = "Variable Importance for At-Vessel Mortality") +
  scale_x_discrete(labels = c("Ventilation Method", "Associated Habitat", "Reproductive Mode", "Maximum Size (cm)", "Median Depth (m)"))
v1

v2 <- prm_fit %>%
  extract_fit_parsnip() %>%
  vip() +
  theme_bw(base_size = 14) +
  labs(y = "Variable Importance for Post-Release Mortality") +
  scale_x_discrete(labels = c("Ventilation Method", "Reproductive Mode", "Associated Habitat", "Maximum Size (cm)"))
v2
```

### Predict additional shark species

```{r}
# filter out rays and missing values
new_dat <- read_csv(here::here("data", "predict_data.csv")) %>%
  filter(order != "Rajiformes") %>%
  filter(order != "Myliobatiformes") %>%
  filter(order != "Torpediniformes") %>%
  # filter(!grepl("batidae", family)) %>%
  filter(family != "Dasyatidae") %>%
  filter(family != "Dasyatidae") %>%
  # filter(family != "Glaucostegidae") %>% # I'd remove this filter
  # filter(family != "Pristidae") %>% # I'd remove this filter
  filter(family != "Trygonorrhinidae") %>%
  filter(!is.na(median_depth)) %>%
  filter(!is.na(ventilation_method)) %>%
  filter(!is.na(reproductive_mode)) %>%
  filter(!is.na(habitat_associated)) %>%
  filter(!is.na(max_size_cm)) %>%
  mutate(ventilation_method = as.factor(ventilation_method)) %>%
  mutate(habitat_associated = as.factor(habitat_associated)) %>%
  mutate(reproductive_mode = as.factor(reproductive_mode)) %>%
  filter(reproductive_mode != "histotrophy")

# predict AVM
predict_data <- list(avm_train, avm_test, new_dat) %>%
  reduce(full_join) %>%
  rename(avm_mort = mortality_prop) %>%
  select(-estimate_type, -method) %>%
  full_join(elasmo_prm) %>%
  rename(prm_mort = mortality_prop) %>%
  select(-estimate_type, -method) %>%
  distinct() %>%
  mutate(habitat_associated = case_when(
    habitat_associated == "pelagic-neritic" ~ "benthopelagic",
    habitat_associated == "pelagic-oceanic" ~ "pelagic",
    TRUE ~ habitat_associated
  )) %>%
  mutate(habitat_associated = as.factor(habitat_associated))

predict_avm <- predict(object = avm_fit, new_data = predict_data) %>% # predict the training set
  bind_cols(predict_data) %>% # bind training set column to prediction
  rename(avm_pred = .pred)

# predict PRM
predict_avm_prm <- predict(object = prm_fit, new_data = predict_avm) %>% # predict the training set
  bind_cols(predict_avm) %>% # bind training set column to prediction
  rename(prm_pred = .pred)

# combine predictions
full_prediction_avm <- preds_bind(predict_avm_prm, avm_fit) %>%
  rename(
    avm_75 = .pred_75,
    avm_95 = .pred_95,
    avm_05 = .pred_05,
    avm_25 = .pred_25,
    avm_50 = .pred_50
  )

full_predictions <- preds_bind(full_prediction_avm, prm_fit) %>%
  rename(
    prm_75 = .pred_75,
    prm_95 = .pred_95,
    prm_05 = .pred_05,
    prm_25 = .pred_25,
    prm_50 = .pred_50
  ) %>%
  filter(family != "Dasyatidae")

write_csv(full_predictions, here::here("data", "full_model_predictions.csv"))
```
 
## Figures for paper

```{r}
plot1 <- (p1 + v1 + q1) / (p2 + v2 + q2) + plot_annotation(tag_levels = "A")

ggsave(plot1, file = paste0("fig2.pdf"), path = here::here("figs"), height = 12, width = 18)
```
