---
title: "random_forest"
author: "Leonardo Feitosa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
library(tidyverse)
library(tidymodels)
library(here)
library(randomForest)
library(ranger)
library(googlesheets4)
library(naniar)
library(ggpmisc)
library(vip)

set.seed(42)
```

```{r}
prm_elasmo <- read_csv(here("data", "prm_hypoxia.csv")) %>% 
  filter(!grepl("Rajidae", family)) %>%
  filter(family != "Dasyatidae") %>%
  filter(family != "Pristidae") %>%
  filter(family != "Rhinidae") %>%
  filter(family != "Rhinobatidae") %>%
  filter(family != "Myliobatidae") %>%
  mutate(habitat = case_when(
    habitat_associated == "pelagic" ~ "pelagic",
    habitat_associated == "pelagic-oceanic" ~ "pelagic",
    habitat_associated == "reef-associated" ~ "pelagic",
    habitat_associated == "bathypelagic" ~ "pelagic",
    habitat_associated == "benthopelagic" ~ "demersal",
    habitat_associated == "demersal" ~ "demersal",
    habitat_associated == "bathydemersal" ~ "demersal",
  )) %>%
    mutate(habitat = case_when(
    scientific_name == "Carcharhinus perezi" ~ "pelagic",
    TRUE ~ habitat
  )) %>%
  mutate(ventilation_method = case_when(
    scientific_name == "Ginglymostoma cirratum" ~ "stationary",
    TRUE ~ ventilation_method
  )) %>%
  mutate(reproductive_mode = case_when(
    scientific_name == "Hexanchus griseus" ~ "lecithotrophic viviparity",
    scientific_name == "Hexanchus nakamurai" ~ "lecithotrophic viviparity",
    scientific_name == "Ginglymostoma cirratum" ~ "lecithotrophic oviparity",
    TRUE ~ reproductive_mode
  )) %>%
  group_by(scientific_name) %>%
  filter(!grepl("sp", scientific_name)) %>% 
  mutate(median_depth = mean(median_depth)) %>%
  mutate(max_size_cm = mean(max_size_cm)) %>%
  distinct()
```

```{r}
# create data subset with variables that will be used in the random forest model
prm_elasmo_subset <- prm_elasmo %>%
  select(
    scientific_name, gear_class, habitat, estimate_type, estimate, sample_size,
    ventilation_method, max_size_cm, median_depth, reproductive_mode, family, ac
  ) %>%
  mutate_if(is.character, as.factor) %>% 
  distinct() %>% 
  filter(sample_size > 5)

# Create subset of the whole dataset with at-vessel mortality and calculate the weighted averages per species and gear
elasmo_trawl <- prm_elasmo_subset %>%
  filter(estimate_type == "at-vessel mortality") %>%
  filter(gear_class == "trawl") %>%
  group_by(scientific_name, gear_class) %>%
  mutate(mortality_prop = weighted.mean(estimate, sample_size)) %>%
  select(ventilation_method, median_depth, max_size_cm, habitat, reproductive_mode, mortality_prop, family, ac) %>%
  distinct() %>%
  drop_na()

# Create subset of the whole dataset with post-release mortality and calculate the weighted averages per species and gear
elasmo_gillnet <- prm_elasmo_subset %>%
  filter(estimate_type == "at-vessel mortality") %>%
  filter(gear_class == "gillnet") %>%
  group_by(scientific_name, gear_class) %>%
  mutate(mortality_prop = weighted.mean(estimate, sample_size)) %>%
  select(ventilation_method, median_depth, max_size_cm, habitat, reproductive_mode, mortality_prop, family, ac) %>%
  distinct() %>%
  drop_na()
```


```{r}
# split the data into training and test set

trawl_folds <- vfold_cv(elasmo_trawl, repeats = 10, v = 5)

gillnet_folds <- vfold_cv(elasmo_gillnet, repeats = 10, v = 5)
```


## Overview

This model aims to predict the likelihood of mortality per taxonomic family between sharks and rays using several species-specific bioecological features such as ventilation mode, median temperature, reproductive mode, median depth, as well as technological features such as the fishing gear each species was captured with.

```{r}
# Create the recipes for each estimate
trawl_recipe <- recipe(mortality_prop ~ ventilation_method + median_depth + max_size_cm + reproductive_mode, data = elasmo_trawl) %>%
  step_normalize(all_numeric_predictors())

gillnet_recipe <- recipe(mortality_prop ~ median_depth + max_size_cm + reproductive_mode, data = elasmo_gillnet) %>%
  step_normalize(all_numeric_predictors())
```

```{r}
# create models
trawl_rf <- rand_forest(trees = tune(), mtry = tune(), min_n = tune(), mode = "regression") %>%
  set_engine("ranger", verbose = TRUE, oob.error = TRUE, quantreg = TRUE, importance = "impurity")

gillnet_rf <- rand_forest(trees = tune(), mtry = tune(), min_n = tune(), mode = "regression") %>%
  set_engine("ranger", verbose = TRUE, oob.error = TRUE, quantreg = TRUE, importance = "impurity")
```

```{r}
# create workflows
trawl_wflw <- workflow() %>%
  add_model(trawl_rf) %>%
  add_recipe(trawl_recipe)

gillnet_wflw <- workflow() %>%
  add_model(gillnet_rf) %>%
  add_recipe(gillnet_recipe)
```

### Cross Validation

Cross validation will be used to choose the best hyperparameters for use in the model: 

* Number of trees
* Number of predictors in each sample
* Minimum data points in each node required for node split

```{r,include=FALSE}
# trawl_tune <- trawl_wflw %>% # use cross validation to tune parameters
#   tune_grid(
#     resamples = trawl_folds, # add cv folds
#     grid = expand.grid(mtry = seq(1, 5, 1), trees = seq(100, 2000, 100), min_n = seq(1, 9, 2))
#   ) # use grid of parameters to test
```

```{r,include=FALSE}
# gillnet_tune <- gillnet_wflw %>% # use cross validation to tune parameters
#   tune_grid(
#     resamples = gillnet_folds, # add cv folds
#     grid = expand.grid(mtry = seq(1, 5, 1), trees = seq(100, 2000, 100), min_n = seq(1, 19, 5))
#   ) # use grid of parameters to test
```

Below displays the results of cross validation tuning

```{r}
# autoplot(trawl_tune) + # plot results of tuning
#   theme_bw() # change theme
```

```{r}
# select_best(trawl_tune, metric = "rmse", n = 1) # get the best model
```

```{r, include=FALSE}
# trawl_final <- finalize_workflow(trawl_wflw, select_best(trawl_tune, metric = "rmse")) # finalize the workflow with the best model
#
# saveRDS(trawl_final, here::here("data", "trawl_final.rds"))

trawl_final <- readRDS(here::here("data", "trawl_final.rds"))
```

```{r, include=FALSE}
trawl_fit <- fit(trawl_final, elasmo_trawl) # fit the data to the training data
```

```{r, include=FALSE}
trawl_train_predict <- predict(object = trawl_fit, new_data = elasmo_trawl) %>% # predict the training set
  bind_cols(elasmo_trawl) # bind training set column to prediction
```

```{r}
trawl_train_metrics <- trawl_train_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
trawl_train_metrics
```

Below displays the results of cross validation tuning

```{r}
# autoplot(gillnet_tune) + # plot results of tuning
#   theme_bw() # change theme
```

```{r}
# select_best(gillnet_tune, metric = "rmse", n = 1) # get the best model
```

```{r, include=FALSE}
# gillnet_final <- finalize_workflow(gillnet_wflw, select_best(gillnet_tune, metric = "rmse")) # finalize the workflow with the best model
#
# saveRDS(gillnet_final, here::here("data", "gillnet_final.rds"))

gillnet_final <- readRDS(here::here("data", "gillnet_final.rds"))
```

```{r, include=FALSE}
gillnet_fit <- fit(gillnet_final, elasmo_gillnet) # fit the data to the training data
```

```{r, include=FALSE}
gillnet_train_predict <- predict(object = gillnet_fit, new_data = elasmo_gillnet) %>% # predict the training set
  bind_cols(elasmo_gillnet) # bind training set column to prediction
```

```{r}
gill_train_metrics <- gillnet_train_predict %>%
  metrics(mortality_prop, .pred) # get testing data metrics
gill_train_metrics
```

```{r}
my_controls <- control_resamples(
  verbose = TRUE,
  save_pred = TRUE,
  extract = function(x) {
    model <- extract_fit_engine(x)
    model$prediction.error
  }
) # set to get prediction and out of bag errors

trawl_cv2 <- trawl_final %>%
  fit_resamples(trawl_folds, control = my_controls) # fit the folds to the final model

gillnet_cv2 <- gillnet_final %>%
  fit_resamples(gillnet_folds, control = my_controls) # fit the folds to the final model
```

```{r}
collect_metrics(trawl_cv2)

collect_metrics(gillnet_cv2)
```

## Quantile regression

```{r}
preds_bind <- function(data_fit, rf_fit) {
  predict(
    rf_fit$fit$fit$fit,
    workflows::extract_recipe(rf_fit) %>% bake(data_fit),
    type = "quantiles",
    quantiles = c(.05, .25, .5, .75, .95)
  ) %>%
    with(predictions) %>%
    as_tibble() %>%
    set_names(paste0(".pred", c("_05", "_25", "_50", "_75", "_95"))) %>%
    bind_cols(data_fit)
} # function to preform and extract quantile estimates from random forest

trawl_quant_test <- preds_bind(trawl_train_predict, trawl_fit) %>%
  mutate(
    range_50 = .pred_75 - .pred_25,
    range_95 = .pred_95 - .pred_05
  )

ggplot(trawl_quant_test) +
  geom_boxplot(aes(range_50)) +
  theme_bw()

write_csv(trawl_quant_test, here::here("data", "trawl_model_predictions.csv"))
```

```{r}
gillnet_quant_test <- preds_bind(gillnet_train_predict, gillnet_fit) %>%
  mutate(
    range_50 = .pred_75 - .pred_25,
    range_95 = .pred_95 - .pred_05
  )

ggplot(gillnet_quant_test) +
  geom_boxplot(aes(range_50)) +
  theme_bw()

write_csv(gillnet_quant_test, here::here("data", "gillnet_model_predictions.csv"))
```

## trawl model - real values vs predicted values

```{r}
p1 <- ggplot(trawl_quant_test, aes(x = mortality_prop, y = .pred)) + # plot ln of real versus ln of predicted
  geom_point() +
  stat_poly_line() +
  stat_poly_eq(use_label("eq")) +
  stat_poly_eq(label.y = 0.9) +
  theme_bw(base_size = 14) +
  labs(
    x = "",
    y = "Predicted At-Vessel Mortality Rate"
  )
p1

summary(lm(.pred ~ mortality_prop, data = trawl_quant_test))
```

## gillnet model - real values vs predicted values

```{r}
p2 <- ggplot(gillnet_train_predict, aes(x = mortality_prop, y = .pred)) + # plot ln of real versus ln of predicted
  geom_point() +
  stat_poly_line() +
  stat_poly_eq(use_label("eq")) +
  stat_poly_eq(label.y = 0.9) +
  theme_bw(base_size = 14) +
  labs(
    x = "Observed At-Vessel Mortality Rate",
    y = "Predicted At-vessel Mortality Rate"
  )
p2

summary(lm(.pred ~ mortality_prop, data = gillnet_train_predict))
```

### Variable importance for the at-vessel mortality and post-release mortality RF models
```{r}
v1 <- trawl_fit %>%
  extract_fit_parsnip() %>%
  vip() +
  theme_bw(base_size = 14) +
  labs(y = "") +
  scale_x_discrete(labels = c("Ventilation Method", "Reproductive Mode", "Median Depth (m)", "Maximum Size (cm)"))
v1

v2 <- gillnet_fit %>%
  extract_fit_parsnip() %>%
  vip() +
  theme_bw(base_size = 14) +
  labs(y = "Variable Importance for At-Vessel Mortality") +
  scale_x_discrete(labels = c("Reproductive Mode", "Maximum Size (cm)", "Median Depth (m)"))
v2
```

## Figures


```{r}
plot1 <- p1 + v1 + p2 + v2 + plot_annotation(tag_levels = "A")

plot1

ggsave(plot1, file = paste0("figS2.pdf"), path = here::here("figs", "supp"), height = 10, width = 15)
```
